---
title: "Mini Project 1 Initial Submission"
author: "Ruchik Rohit Dama,Aniket Kale, Shashank Kumar, Sumanth Gopalkrishna"
date: "9/29/2021"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readxl)
library(ggplot2)
library(tidycensus)
library(stringr)
library(tidyverse)
library(gridExtra)
census_api_key("b58a893ab08bf54aa71a88588839c192abdb915a")
```
## Question 1

### Load the data

```{r}
cpi_data=read_xls("State_and_US_SA.xls", range = 'A6:BA546', col_names = TRUE)
cpi_data1=read_xls("State_and_US_SA.xls", range = 'A6:BA546', col_names = TRUE)
cpi=read.csv(file = 'cpi.csv')
```


```{r}
tail(cpi)
```




```{r}
col_2019=subset(cpi, cpi$Year == 2019)
vals=col_2019$Value
cpi$cpi_2019 <- rep(vals, times = 73)
#head(cpi)
```

### Removed the data prior to 1975.
```{r}
cpi=cpi[which(cpi$Year>1974),]
#head(cpi)
```

### Splitted the month column and stored it in the Month and Year Columns


```{r}
cpi_temp=str_split_fixed(cpi_data$Month, "M", 2)
cpi_data1$Year=as.numeric(cpi_temp[,1])
cpi_data1$Month=as.numeric(cpi_temp[,2])
cpi_data$Year=as.numeric(cpi_temp[,1])
cpi_data$Month=as.numeric(cpi_temp[,2])
#head(cpi_data)
```

### Merging cpi_data and cpi dateset
```{r}
cpi_data = merge(cpi_data, cpi, by = c("Year"),inplace=TRUE)
```


### Renaming United States seasonally adjusted to usprice 
```{r}
colnames(cpi_data)[colnames(cpi_data) == "United States seasonally adjusted"] = "US"
head(cpi_data)
```




```{r}
cpi_data[,3:53] = (cpi_data$cpi_2019*cpi_data[,3:53]/cpi_data$Value) 
head(cpi_data)
```
```{r}
cpi_data$US = (cpi_data$cpi_2019*cpi_data$US/cpi_data$Value) 
head(cpi_data)
```
### Question 1.1:- How  have  house  prices  in  the  U.S  changed  since  1975,  afteradjusting for inflation (i.e. relative to the CPI?) 
### Plloting the graph between years v/s change in value before and after adjusting inflation.
```{r}
#ggplot(cpi_data1,aes(x = Year , y = `United States seasonally adjusted`))+geom_line()+labs(title = "Year v/s change in value",x=" Year",y="US seasonally Adjusted")
```

```{r}
ggplot(cpi_data,aes(x = Year , y = US))+geom_line()+labs(title = "Year v/s change in value",x=" Year",y="US seasonally Adjusted")  
```

SOME EXPLANATION HERE. 
### Loading  the statecodes text file

```{r}
state_codes_df = read.delim(file = "statecodes.txt", header = FALSE, sep = "", dec = ".")
tail(cpi_data)
```



```{r}
state_codes=state_codes_df$V2
head(cpi_data)
```

### Pivoting the data from wide to long.
```{r}
state_data=cpi_data[1:54]
state_data1=state_data %>% 
 pivot_longer(
   cols =state_codes,
   names_to = "variable",
   values_to = "value",
   values_drop_na = TRUE
 )
```
```{r}
head(state_data)
```


### Finding the mean of values.
```{r}
state_mean=aggregate(value~variable + Year,data=state_data1,mean)
```
### Question 1.2:- How have changes in prices varied by state?
### Plotting graphs for each state


```{r}
west_states = c("WA", "MT", "OR", "ID", "WY", "CA", "NV", "UT", "CO", "AZ", "NM", "HI", "AK")
midwest_states = c("ND", "MN", "SD", "NE", "IA", "KS", "MO", "WI", "MI", "IL", "IN", "OH")
northeast_states = c("PA", "NJ", "NY", "CT", "RI", "MA", "VT", "NH", "ME")
south_states = c("OK", "TX", "AR", "LA", "MS", "AL", "TN", "KY", "GA", "FL", "SC", "NC", "WV", "VA", "DC", "MD", "DE")
west_data = state_mean[state_mean$variable %in% west_states,]
midwest_data = state_mean[state_mean$variable %in% midwest_states,]
northeast_data = state_mean[state_mean$variable %in% northeast_states,]
south_data = state_mean[state_mean$variable %in% south_states,]
```

```{r}
head(west_data)
```


```{r}
west_mean=aggregate(value~variable+Year,data=west_data,mean)
midwest_mean=aggregate(value~variable+Year,data=midwest_data,mean)
northeast_mean=aggregate(value~variable+Year,data=northeast_data,mean)
south_mean=aggregate(value~variable+Year,data=south_data,mean)
```

```{r}
west_plot <- ggplot(west_mean, aes(x=Year, y = value)) + geom_line() 
west_plot + facet_wrap(~variable, ncol = 6,scales = "free_x")  + theme_minimal()+theme(axis.text.x = element_text(angle = 60, vjust = 0.5, hjust=1))
```


```{r}
midwest_plot <- ggplot(midwest_mean, aes(x=Year, y = value)) + geom_line() 
midwest_plot + facet_wrap(~variable, ncol = 6)  + theme_minimal()+theme(axis.text.x = element_text(angle = 60, vjust = 0.5, hjust=1))
```

```{r}
northeast_plot <- ggplot(northeast_mean, aes(x=Year, y = value)) + geom_line() 
northeast_plot + facet_wrap(~variable, ncol = 6)  + theme_minimal()+theme(axis.text.x = element_text(angle = 60, vjust = 0.5, hjust=1))
```


```{r}
south_plot <- ggplot(south_mean, aes(x=Year, y = value)) + geom_line() 
south_plot + facet_wrap(~variable, ncol = 6)  + theme_minimal()+theme(axis.text.x = element_text(angle = 60, vjust = 0.5, hjust=1))
```


The state wise plots for house prices vs timeline are as above. They pretty much follow similar behaviour as
we saw in the previous plot. Although, we do see certain states which have a sudden rise like DC i.e. District of
Columbia. 

### QUestion 1.3:- Which states have seen the biggest increases in real house prices, and which have seen the biggest decreases
### States with Biggest Increases
### Plotting graphs of top 4 states with biggest increases
```{r}
#state_plot <- ggplot(state_mean, aes(x=Year, y = value)) + geom_line() 
#state_plot + facet_wrap(~variable, ncol = 7)  + theme_minimal()
dc_df = subset(state_mean, state_mean$variable == "DC")
cal_df = subset(state_mean, state_mean$variable == "CA")
hi_df = subset(state_mean, state_mean$variable == "HI")
wa_df = subset(state_mean, state_mean$variable == "WA")
g1 = ggplot(dc_df, aes(x = Year, y = value)) + geom_line() + ggtitle("DISTRICT OF COLUMBIA") 
g2 = ggplot(cal_df, aes(x = Year, y = value)) + geom_line() + ggtitle("CALIFORNIA")
g3 = ggplot(hi_df, aes(x = Year, y = value)) + geom_line() + ggtitle("HAWAII")
g4 = ggplot(wa_df, aes(x = Year, y = value)) + geom_line() + ggtitle("WASHINGTON")
grid.arrange(g1,g2,g3,g4, nrow = 2)
```
### States with Biggest Decreases
### Plotting graphs of top 4 states with biggest decreases
```{r}
ar_df = subset(state_mean, state_mean$variable == "AR")
oh_df = subset(state_mean, state_mean$variable == "OH")
ok_df = subset(state_mean, state_mean$variable == "OK")
wv_df = subset(state_mean, state_mean$variable == "WV")
g1 = ggplot(ar_df, aes(x = Year, y = value)) + geom_line() + ggtitle("ARKANSAS") 
g2 = ggplot(oh_df, aes(x = Year, y = value)) + geom_line() + ggtitle("OHIO")
g3 = ggplot(ok_df, aes(x = Year, y = value)) + geom_line() + ggtitle("OKLOHOMA")
g4 = ggplot(wv_df, aes(x = Year, y = value)) + geom_line() + ggtitle("WEST VIRGINIA")
grid.arrange(g1,g2,g3,g4, nrow = 2)
```

### Question 1.4:- Have changes in prices within each state mostly followed the same basicpattern,  and are there outliers to that pattern

the house prices within each state have mostly followed the basic pattern of rises and falls cycle. And we
have observed outliers where they defy that pattern by showing a continuous drop cycle or rise cycle. The
outlier states are West Virginia, Arkansas, District of Columbia, California, Washington. There
might be more as well. 

### Question 1.5:- Do the typical patterns vary between thefour regions (Northeast, Midwest, South, and West)?

### Grpah for 4 regions namely:-
### Plotting graphs of west, midwest, northeast and south states.


```{r}
west_mean1=aggregate(value~Year,data=west_data,mean)
midwest_mean1=aggregate(value~Year,data=midwest_data,mean)
northeast_mean1=aggregate(value~Year,data=northeast_data,mean)
south_mean1=aggregate(value~Year,data=south_data,mean)
#south_mean
```

```{r}
g1 = ggplot(west_mean1, aes(x = Year, y = value)) + geom_line() + ggtitle("WEST") 
g2 = ggplot(midwest_mean1, aes(x = Year, y = value)) + geom_line() + ggtitle("MIDWEST")
g3 = ggplot(northeast_mean1, aes(x = Year, y = value)) + geom_line() + ggtitle("NORTHEAST")
g4 = ggplot(south_mean1, aes(x = Year, y = value)) + geom_line() + ggtitle("SOUTH")
grid.arrange(g1,g2,g3,g4, nrow = 2)
```
The regions also pretty much follow the same trend as the overall house prices.
Northeast region has a gradual rise and so does the West region. The period 1980-1990 was a consistently
low patch for all except the Northeast region. 

### Part 2:

```{r}
# import libararies

library(tidyverse)
library(broom)
library(readxl)
library(tidycensus)
library(arm)
library(MASS)
#census_api_key("10f0d49b7a43755878e4706e12c9b552b96428f7")

```

```{r}
#read data files
cpi = read.csv("cpi.csv")
pop = read.csv("population.csv")
areas = read.csv("https://raw.githubusercontent.com/jakevdp/data-USstates/master/state-areas.csv", stringsAsFactors=FALSE, fileEncoding="latin1")

scode = read.delim("statecodes.txt", header=F, sep=" ")
hpi = read_excel("State_and_US_SA.xls", skip=5)
```

```{r}
# loading tidycensus data
# census20 =  get_decennial(geography = "state", 
#                        variables = "P001001", 
#                        year = 2020)
# census10 =  get_decennial(geography = "state", 
#                        variables = "P001001", 
#                        year = 2010)
# census00 =  get_decennial(geography = "state", 
#                        variables = "P001001", 
#                        year = 2000)
# census90 = get_decennial(geography = "state", 
#                        variables = "P0010001", 
#                        year = 1990)
# View(census10)
```

```{r}
# head(areas)

# density.00 = merge(
#   x=census00,
#   y=areas,
#   by.x="NAME", by.y="state")
# density.00 = density.00%>%
#   mutate(density = value/area..sq..mi.)
```

```{r}
# Pre-process population df
names(pop)[2] = "state"
names(pop)[5] = "total.1970"
names(pop)[6] = "total.1980"
names(pop)[7] = "total.1990"
names(pop)[8] = "total.2000"
names(pop)[9] = "total.2010"
names(pop)[10] = "total.2008.2012"
pop = pop[,c( "state",
              "total.1970",
              "total.1980",
              "total.1990",
              "total.2000",
              "total.2010",
              "total.2008.2012"
              )]
# head(pop)
```

```{r}
# computing population density for all the states, after merging with areas df
pop.density = merge(
  x=pop,
  y=areas,
  by.x="state", by.y="state")

pop.density = pop.density %>%
  mutate(desity.1970 = total.1970/area..sq..mi.,
        desity.1980 = total.1980/area..sq..mi.,
        desity.1990 = total.1990/area..sq..mi.,
        desity.2000 = total.2000/area..sq..mi.,
        desity.2010 = total.2010/area..sq..mi.,
        desity.2008.2012 = total.2008.2012/area..sq..mi.
)
pop.density = merge(
  x=pop.density%>% mutate(state = toupper(state)),
  y=scode,
  by.x="state", by.y="V3")
# pop.density
```
```{r}



pop.density = dplyr::select(pop.density, -c("V1", "V4", "V5"))
names(pop.density)[length(names(pop.density))] = "ST"
# head(pop.density)
```

```{r}
hpi.dec = hpi %>% filter(
  Month %in% c("1980M12", "1990M12", "2000M12", "2010M12", "2019M12")
)
```


```{r}
scode = scode %>% filter(
  V2 != ""
)
print("Updated rows:")
print(nrow(scode))
```

```{r}
hpi.dec.t = as.data.frame(t(hpi.dec))
names(hpi.dec.t) <- as.matrix(hpi.dec.t[1, ])
hpi.dec.t <- hpi.dec.t[-1, ]
hpi.dec.t[] <- lapply(hpi.dec.t, function(x) type.convert(as.character(x)))
hpi.dec.t <- cbind(ST = rownames(hpi.dec.t), hpi.dec.t)
rownames(hpi.dec.t) <- 1:nrow(hpi.dec.t)
# head(hpi.dec.t)
```
```{r}
 
pop.hpi.merged = merge(
  x=hpi.dec.t,
  y=pop.density,
  by.x="ST", by.y="ST")
names(pop.hpi.merged)[2] = "hpi.1980"
names(pop.hpi.merged)[3] = "hpi.1990"
names(pop.hpi.merged)[4] = "hpi.2000"
names(pop.hpi.merged)[5] = "hpi.2010"
names(pop.hpi.merged)[6] = "hpi.2019"

# head(pop.hpi.merged)
```
```{r}
colors <- c("1980" = "black", "1990" = "blue", "2000" = "darkgreen", "2010" = "orange")

pop.hpi.merged %>%
  ggplot(aes(x=hpi.1980, y =desity.1980, color="1980")) +geom_point() +
  geom_point(data = pop.hpi.merged, aes(x=hpi.1990, y =desity.1990, col = "1990"))+
  geom_point(data = pop.hpi.merged, aes(x=hpi.2000, y =desity.2000, col = "2000"))+
  geom_point(data = pop.hpi.merged, aes(x=hpi.2010, y =desity.2010, col = "2010"))+
  labs(y = "HPI",
         x = "Population Density",
         color = "Legend")+
    scale_color_manual(values = colors)+
  geom_smooth(method = "lm" ,se=F)+
  geom_smooth(method = "lm" ,se=F,data = pop.hpi.merged, aes(x=hpi.1990, y =desity.1990, col = "1990"))+
  geom_smooth(method = "lm" ,se=F,data = pop.hpi.merged, aes(x=hpi.2000, y =desity.2000, col = "2000"))+
  geom_smooth(method = "lm" ,se=F,data = pop.hpi.merged, aes(x=hpi.2010, y =desity.2010, col = "2010"))+
  facet_wrap(~state)
```
```{r}
colors <- c("1980" = "black", "1990" = "blue", "2000" = "darkgreen", "2010" = "orange")

pop.hpi.merged %>%
  ggplot(aes(y=hpi.1980, x =desity.1980, color="1980")) +
  geom_point() +
  geom_point(data = pop.hpi.merged, aes(y=hpi.1990, x =desity.1990, col = "1990"))+
  geom_point(data = pop.hpi.merged, aes(y=hpi.2000, x =desity.2000, col = "2000"))+
  geom_point(data = pop.hpi.merged, aes(y=hpi.2010, x =desity.2010, col = "2010"))+
  labs(y = "HPI",
         x = "Population Density",
         color = "Legend")+
    scale_color_manual(values = colors)+
  geom_smooth(method = "lm" ,se=F)+
  geom_smooth(method = "lm" ,se=F,data = pop.hpi.merged, aes(y=hpi.1990, x =desity.1990, col = "1990"))+
  geom_smooth(method = "lm" ,se=F,data = pop.hpi.merged, aes(y=hpi.2000, x =desity.2000, col = "2000"))+
  geom_smooth(method = "lm" ,se=F,data = pop.hpi.merged, aes(y=hpi.2010, x =desity.2010, col = "2010"))
```

```{r}
colors <- c("1980" = "black", "1990" = "blue", "2000" = "darkgreen", "2010" = "orange")

pop.hpi.merged %>%
  ggplot(aes(x=hpi.1980, y =desity.1980, color="1980")) +
  geom_point() +
  geom_point(data = pop.hpi.merged, aes(x=hpi.1990, y =desity.1990, col = "1990"))+
  geom_point(data = pop.hpi.merged, aes(x=hpi.2000, y =desity.2000, col = "2000"))+
  geom_point(data = pop.hpi.merged, aes(x=hpi.2010, y =desity.2010, col = "2010"))+
  labs(x = "HPI",
         y = "Population Density",
         color = "Legend")+
    scale_color_manual(values = colors)+
  geom_smooth(method = "lm" ,se=F)+
  geom_smooth(method = "lm" ,se=F,data = pop.hpi.merged, aes(x=hpi.1990, y =desity.1990, col = "1990"))+
  geom_smooth(method = "lm" ,se=F,data = pop.hpi.merged, aes(x=hpi.2000, y =desity.2000, col = "2000"))+
  geom_smooth(method = "lm" ,se=F,data = pop.hpi.merged, aes(x=hpi.2010, y =desity.2010, col = "2010"))
```

#### Outliers:
* There are two influencial outliers for every year.
* Instead of completely removing outliers, we can try to down weigh them. Because these points represent two states and they just cannot be eliminated.

```{r}

# finding outliers
pop.hpi.merged %>%
  filter(desity.1980>500 | desity.1990>500 | desity.2000>500 | desity.2010>500)
```

method = rlm, se=F, col = "orange",
              method.args=list(psi = psi.bisquare)
              

```{r}

# after using rlm
colors <- c("1980" = "black", "1990" = "blue", "2000" = "darkgreen", "2010" = "orange")

pop.hpi.merged %>%
  ggplot(aes(x=hpi.1980, y =desity.1980, color="1980")) +
  geom_point() +
  geom_point(data = pop.hpi.merged, aes(x=hpi.1990, y =desity.1990, col = "1990"))+
  geom_point(data = pop.hpi.merged, aes(x=hpi.2000, y =desity.2000, col = "2000"))+
  geom_point(data = pop.hpi.merged, aes(x=hpi.2010, y =desity.2010, col = "2010"))+
  labs(x = "HPI",
         y = "Population Density",
         color = "Legend")+
    scale_color_manual(values = colors)+
  geom_smooth(method = rlm ,se=F)+
  geom_smooth(method = rlm,method.args = list(psi.bisquare) ,se=F,data = pop.hpi.merged, aes(x=hpi.1990, y =desity.1990, col = "1990"))+
  geom_smooth(method = rlm,method.args = list(psi.bisquare) ,se=F,data = pop.hpi.merged, aes(x=hpi.2000, y =desity.2000, col = "2000"))+
  geom_smooth(method = rlm,method.args = list(psi.bisquare) ,se=F,data = pop.hpi.merged, aes(x=hpi.2010, y =desity.2010, col = "2010"))

```


### QUESTION 3:-

### Loading the Data


```{r}
population_2018 = get_acs(geography = "state", 
              variables = c(Total = "B01003_001"), 
              year = 2018)

```

```{r}
head(population_2018)
```




```{r}
population=read.csv(file = 'population.csv')
#(population)
```
```{r}
colnames(state_codes_df)[colnames(state_codes_df) == "V2"] <- "variable"

colnames(state_codes_df)[colnames(state_codes_df) == "V3"] <- "NAME"
state_codes_df$NAME <- toupper(state_codes_df$NAME)
```


```{r}
#head(state_codes_df)
```

### Storing the data for 2018, 2010, 2000, 1990 year.

```{r}
state_mean <- merge(state_mean, state_codes_df, by = "variable")
#head(state_mean)
```
```{r}
pop_10 <- subset(state_mean, state_mean$Year == 2010)
pop_90 <- subset(state_mean, state_mean$Year == 1990)
pop_00 <- subset(state_mean, state_mean$Year == 2000)
pop_18 <- subset(state_mean, state_mean$Year == 2018)
```

```{r}

colnames(population)[colnames(population) == "NHGIS.Integrated.State.Name"] <- "NAME"
#head(population)
population$NAME <- toupper(population$NAME)
```



```{r}
population_1990=population[c(2,7)]
population_2000=population[c(2,8)]
population_2010=population[c(2,9)]
#population_2018=population_2018[c(2,4)]
#head(population_1990)
```
```{r}

population_2018$NAME = toupper(population_2018$NAME)
#head(population_2018)
```
```{r}
#head(pop_10)
```


```{r}
colnames(population_1990)[colnames(population_1990) == "X1990..Persons..Total"] <- "Total_1990"
colnames(population_2000)[colnames(population_2000) == "X2000..Persons..Total"] <- "Total_2000"
colnames(population_2010)[colnames(population_2010) == "X2010..Persons..Total"] <- "Total_2010"
colnames(population_2018)[colnames(population_2018) == "estimate"] <- "Total_2018"
```

### Merging datasets to get a common dataset for each year.

```{r}
data_10 <- merge(pop_10, population_2010, by = "NAME")
data_00 <- merge(pop_00, population_2000, by = "NAME")
data_90 <- merge(pop_90, population_1990, by = "NAME")
data_18 <- merge(pop_18, population_2018, by = "NAME")
```

```{r}
colnames(pop_90)[colnames(pop_90) == "V3"] <- "NAME"
colnames(pop_00)[colnames(pop_00) == "V3"] <- "NAME"
colnames(pop_10)[colnames(pop_10) == "V3"] <- "NAME"

```

```{r}
data_18=data_18[c("NAME","value","Total_2018")]
data_00=data_00[c("NAME","value","Total_2000")]
data_10=data_10[c("NAME","value","Total_2010")]
data_90=data_90[c("NAME","value","Total_1990")]

```

### Finding change in house prices and change in population.
```{r}
data_10$pdiff <- data_10$Total_2010 - data_00$Total_2000
data_18$pdiff <- data_18$Total_2018 - data_10$Total_2010
data_00$pdiff <- data_00$Total_2000 - data_90$Total_1990
data_10$diff <- data_10$value - data_00$value
data_18$diff <- data_18$value - data_10$value
data_00$diff <- data_00$value - data_90$value
```

```{r}
#head(data_00)
```

### Plotting the graphs for change in house price and population

```{r}
g1 = ggplot(data_18, aes(x = pdiff, y = diff)) + geom_point() + xlab("Difference in population") + ylab("Difference in house price") + ggtitle(" 2018 minus 2010")+geom_smooth()

g2 = ggplot(data_10, aes(x = pdiff, y = diff)) + geom_point() + xlab("Difference in population") + ylab("Difference in house price") + ggtitle("2010 minus 2000")+geom_smooth()

g3 = ggplot(data_00, aes(x = pdiff, y = diff)) + geom_point() + xlab("Difference in population") + ylab("Difference in house price") + ggtitle("2000 minus 1990")+geom_smooth()

grid.arrange(g1,g2,g3, nrow = 1)
```
 
```{r}
data_18 = merge(data_18, state_codes_df, by = "NAME")
data_10 = merge(data_10, state_codes_df, by = "NAME")
data_00 = merge(data_00, state_codes_df, by = "NAME")
data_90 = merge(data_90, state_codes_df, by = "NAME")


```

```{r}
west_df = data_18[data_18$variable %in% west_states,]
midwest_df = data_18[data_18$variable %in% midwest_states,]
northeast_df = data_18[data_18$variable %in% northeast_states,]
south_df = data_18[data_18$variable %in% south_states,]
```


### plotting the graphs for west, midwest, northeast, south for change in house prices and population for every year.

```{r}
g1 = ggplot(west_df, aes(x = pdiff, y = diff)) + geom_point() + ggtitle("WEST") +geom_smooth()
g2 = ggplot(midwest_df, aes(x = pdiff, y = diff)) + geom_point() + ggtitle("MIDWEST")+geom_smooth()
g3 = ggplot(northeast_df, aes(x = pdiff, y = diff)) + geom_point() + ggtitle("NORTHEAST")+geom_smooth()
g4 = ggplot(south_df, aes(x = pdiff, y = diff)) + geom_point() + ggtitle("SOUTH")+geom_smooth()
grid.arrange(g1,g2,g3,g4, nrow = 2)
```
```{r}
west_df = data_10[data_10$variable %in% west_states,]
midwest_df = data_10[data_10$variable %in% midwest_states,]
northeast_df = data_10[data_10$variable %in% northeast_states,]
south_df = data_10[data_10$variable %in% south_states,]
```

```{r}
g1 = ggplot(west_df, aes(x = pdiff, y = diff)) + geom_point() + ggtitle("WEST") +geom_smooth()
g2 = ggplot(midwest_df, aes(x = pdiff, y = diff)) + geom_point() + ggtitle("MIDWEST")+geom_smooth()
g3 = ggplot(northeast_df, aes(x = pdiff, y = diff)) + geom_point() + ggtitle("NORTHEAST")+geom_smooth()
g4 = ggplot(south_df, aes(x = pdiff, y = diff)) + geom_point() + ggtitle("SOUTH")+geom_smooth()
grid.arrange(g1,g2,g3,g4, nrow = 2)
```

```{r}
west_df = data_00[data_10$variable %in% west_states,]
midwest_df = data_00[data_00$variable %in% midwest_states,]
northeast_df = data_00[data_00$variable %in% northeast_states,]
south_df = data_00[data_00$variable %in% south_states,]
```


```{r}
g1 = ggplot(west_df, aes(x = pdiff, y = diff)) + geom_point() + ggtitle("WEST") +geom_smooth()
g2 = ggplot(midwest_df, aes(x = pdiff, y = diff)) + geom_point() + ggtitle("MIDWEST")+geom_smooth()
g3 = ggplot(northeast_df, aes(x = pdiff, y = diff)) + geom_point() + ggtitle("NORTHEAST")+geom_smooth()
g4 = ggplot(south_df, aes(x = pdiff, y = diff)) + geom_point() + ggtitle("SOUTH")+geom_smooth()
grid.arrange(g1,g2,g3,g4, nrow = 2)
```


```{r}
v17 <- load_variables(2017, "acs5", cache = TRUE)

View(v17)

```


Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.