---
title: "Assignment_4_Final"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
library(mgcv)
library(readr)
library("metR")
```


```{r}
basics=read_tsv("data_basics.tsv", na = "\\N", quote = '')
ratings=read_tsv("data_ratings.tsv", na = "\\N", quote = '')
```

```{r}
head(basics)
head(ratings)
```

#### Merge the two data frames
```{r}
df<- merge(basics,ratings,by='tconst')
head(df)
```


#### Subset the data frame for only movies
```{r}
df<-df[df$titleType=='movie',]
head(df)
```

#### Check shape 
```{r}
dim(df)
```

#### Checking for null values
```{r}
sapply(df, function(x) sum(is.na(x)))
```


#### Considering only the required columns into consideration
```{r}
df<-df[c('startYear','runtimeMinutes','averageRating','numVotes')]
head(df)
```

#### Dropping all those rows with NA values.
```{r}
print('before dropping null values')
dim(df)
df<-na.omit(df)
dim(df)
```

#### check distributions of each independent variable
```{r}
#ggplot(df,aes(x=startYear))+geom_density()+xlab('start year')+ylab('count')+ggtitle('Year distribution')

#ggplot(df,aes(x=runtimeMinutes))+geom_density()+xlab('runtime')+ylab('Count')+xlim(50,200)+ggtitle('runtime distribution')

#ggplot(df,aes(x=averageRating))+geom_density()+xlab('averageRating')+ylab('count')+ggtitle('averageRating distribution')

```

```{r}
summary(df)
```

#### It looks like there are a lot of outliers for the runtimeMinutes. Philosophically,<br>Movies tend to be in the range of 50 minutes (shroter movies) to 200 minutes(longer movies)<br>Dropping all movies with length lesser than 50 and greater than 200 minutes
```{r}
df<-subset(df,df$runtimeMinutes<200 & df$runtimeMinutes>50)
dim(df)
```



### lm model
```{r}
model <- lm(averageRating~startYear+runtimeMinutes,weights=numVotes,data=df)
summary(model)
```

##weighted
```{r}
model <- gam(averageRating~startYear+s(runtimeMinutes),data=df,weights=numVotes,method='REML')
summary(model)
```


```{r}
model.df = data.frame(df,
                   .fitted = fitted.values(model),
                   .resid = residuals(model))
head(model.df)
```

```{r}
merge.grid=expand.grid(runtimeMinutes=50:200,startYear=1896:2021)
merge.predict=predict(model,newdata=merge.grid)
merge.grid.df=data.frame(merge.grid,avgRating=as.vector(merge.predict))
```


```{r}
head(merge.grid.df)
```

```{r}
ggplot(merge.grid.df, aes(x = runtimeMinutes, y=avgRating)) + geom_point()+
geom_smooth(method='gam')+
facet_wrap(~cut_width(merge.grid.df$startYear,10))+
ggtitle('AverageRating vs RuntimeMinutes')+
labs(subtitle = "Cut by Years in Decades")
```

#contour plot
```{r,fig.height=7,fig.width=12}
ggplot(merge.grid.df,aes(x = runtimeMinutes, y = startYear, z =avgRating,fill=avgRating))+geom_raster()+
  scale_fill_distiller(palette = "RdYlBu")+ 
  geom_contour()+
  geom_text_contour()+
  ggtitle('Raster and Counter')
```