---
title: "ProbSet5"
author: "Himani Anil Deshpande"
date: "10/23/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(MASS)
library(arm)
library(broom)
library(grid)
library(png)
library(jpeg)
library(GGally)
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```
# Loading Data

```{r}
nba.data = read.table("curry2015-16.txt", header= 1)


nba.data = nba.data[,c('EVENT_TYPE','SHOT_MADE_FLAG','SHOT_DISTANCE','LOC_X','LOC_Y')]
view(nba.data)
```
## Plot the data to show the location of Curry’s shots using color to distinguish between made
and missed shots, similarly to the picture below but more colorblind-friendly. (You don’t
have to include the picture of the court unless you want to show off.) NB: It should use
coord fixed() since the units are the same for both axes.

```{r}

#https://content.sportslogos.net/logos/6/235/full/141_golden_state-warriors-stadium-2012.png

#background_img = "golden_state-warriors-stadium-2012.png"

#court_background =  rasterGrob(readPNG(background_img), interpolate = TRUE, width=unit(1, "npc"), height=unit(1, "npc"))

ggplot(nba.data, aes(x = LOC_X, y = LOC_Y, color = EVENT_TYPE)) +
  #annotation_custom(court_background, xmin=-Inf, xmax=Inf, ymin=-Inf, ymax=Inf) +
  geom_point() + 
  coord_fixed(ratio = 1) #ratio = 1 means units are the same for both axes.



#background_img = "basketball-court-top-view-vector-illustration-on-black-background-1.jpeg"



#court_background =  rasterGrob(readJPEG(background_img), interpolate = TRUE, width=unit(1, "npc"), height=unit(1, "npc"))

ggplot(nba.data, aes(x = LOC_X, y = LOC_Y, color = EVENT_TYPE)) +
  #annotation_custom(court_background, xmin=-Inf, xmax=Inf, ymin=-Inf, ymax=Inf) +
  geom_point() + 
  coord_fixed(ratio = 1) #ratio = 1 means units are the same for both axes.


```
## Fit a logistic regression to predict whether the shot is made, using the single predictor
SHOT DISTANCE. Draw an appropriate ggplot of the fitted curve and write an equation for
the fit
```{r}
ggpairs(nba.data) #Correlation between SHOT_DISTANCE and LOC_Y is high
#cor(nba.data$SHOT_DISTANCE, nba.data$EVENT_TYPE)
# cor( nba.data$SHOT_MADE_FLAG, nba.data$SHOT_DISTANCE)
# cor(  nba.data$SHOT_DISTANCE, nba.data$SHOT_MADE_FLAG)

nba.model.glm = glm(SHOT_MADE_FLAG ~ SHOT_DISTANCE, data = nba.data, family = "binomial")
nba.model.grid = expand.grid(SHOT_DISTANCE = seq(0,90, 5))

nba.model.pred = predict(nba.model.glm, newdata = nba.model.grid, type = "response")
  nba.model.grid = data.frame(nba.model.grid, prob = as.vector(nba.model.pred))

ggplot(nba.model.grid, aes(x = SHOT_DISTANCE, y = prob))+geom_line()



# 
# 
# model1.grid = data.frame(model.grid, prob = as.vector(model1.pred))
# 
# ggplot(model1.grid, aes(x = , y = prob, group = dist, color = dist))+geom_line()
#   
# 
# ggplot(model1.grid, aes(x = arsenic, y = prob, group = dist, color = dist))+geom_line() +
#   scale_x_log10()
# 
#   
```

```{r}
nba.model.grid = expand.grid(SHOT_DISTANCE = seq(0,90, 5))
head(nba.model.grid)
```


```{r}
load("lattice.RData")
summary(iris)
ggpairs(iris)
```
# Plot the residuals in a way that shows where the logistic regression doesn’t fit the data well.
Describe in some detail how the model is inaccurate.
```{r}
nba.model.glm.df = data.frame(nba.data, .resid = residuals(nba.model.glm) )

ggplot(nba.model.glm.df, aes(x = .resid, y = SHOT_MADE_FLAG, color = EVENT_TYPE))+
         geom_point()
```
#Fit a better model. You could try a different functional form or a model with more predictors
(as long as you use the predictors sensibly.) Your model doesn’t have to be perfect, just
better. Draw a graph that shows how your model differs from the simple logistic regression,
and convince us that your model is better.
```{r}
# nba.model.glm2 = glm(SHOT_MADE_FLAG ~ SHOT_DISTANCE +s(LOC_Y), data = nba.data, family = binomial)
# nba.model.grid2 = expand.grid(SHOT_DISTANCE = seq(0,90, 5))


AIC(glm(SHOT_MADE_FLAG ~ SHOT_DISTANCE, data = nba.data, family = "binomial"))

AIC(glm(SHOT_MADE_FLAG ~ SHOT_DISTANCE +LOC_Y, data = nba.data, family = "binomial"))

AIC(glm(SHOT_MADE_FLAG ~ SHOT_DISTANCE +LOC_Y + LOC_X, data = nba.data, family = "binomial"))


# AIC(glm(SHOT_MADE_FLAG ~ log10(SHOT_DISTANCE), data = nba.data, family = "binomial"))
# 
# AIC(glm(SHOT_MADE_FLAG ~ log10(SHOT_DISTANCE) +LOC_Y, data = nba.data, family = "binomial"))
# 
# AIC(glm(SHOT_MADE_FLAG ~ log10(SHOT_DISTANCE) +LOC_Y + LOC_X, data = nba.data, family = "binomial"))



AIC(glm(SHOT_MADE_FLAG ~ SHOT_DISTANCE* LOC_Y , data = nba.data, family = "binomial"))
AIC(glm(SHOT_MADE_FLAG ~ SHOT_DISTANCE* LOC_Y *LOC_X, data = nba.data, family = "binomial"))

AIC(glm(SHOT_MADE_FLAG ~ SHOT_DISTANCE *LOC_X+LOC_Y, data = nba.data, family = "binomial"))


AIC(glm(SHOT_MADE_FLAG ~ SHOT_DISTANCE+ LOC_X:LOC_Y, data = nba.data, family = "binomial"))

AIC(glm(SHOT_MADE_FLAG ~ SHOT_DISTANCE+ LOC_Y:LOC_X, data = nba.data, family = "binomial"))# better
AIC(glm(SHOT_MADE_FLAG ~ SHOT_DISTANCE* LOC_Y :LOC_X, data = nba.data, family = "binomial"))
```
```{r}
modelX = glm(SHOT_MADE_FLAG ~ SHOT_DISTANCE+ LOC_Y:LOC_X , family = "binomial", data = nba.data)
modelX

```

```{r}
modelX.grid = expand.grid(SHOT_DISTANCE = seq(0,90, 5), LOC_X = -300:300, LOC_Y = -100:800)

modelX.grid$prob =as.vector(predict(modelX, newdata = modelX.grid, type ="response"))
ggplot(modelX.grid, aes(x = SHOT_DISTANCE, y = prob ))+ 
  geom_line()

```






